%%%% Time-stamp: <2013-02-25 10:31:01 vk>

\phantomsection
\addcontentsline{toc}{chapter}{Abstract}
\chapter*{Abstract}
\label{cha:abstract}

Microarchitectural attacks exploit mistakes made by hardware vendors, who rely
heavily on a perfect design. Cache-attacks or attacks like Meltdown and Spectre
target the CPU, where they either exploit a principle which works as designed
and is therefore vulnerable, or abuse bugs introduced by mistakes in the
hardware design or the microcode. Rowhammer makes use of a design flaw
introduced by DRAM vendors, who were forced to lower quality by decreasing the
size and refresh rates of memory cells. With specially crafted memory access
routines it is possible to access DRAM cells fast enough to create interfering
loads between cells, which causes others to change their load, hence switching
their logic state.

Rowhammer attacks use this phenomenon to target memory areas where flips would
benefit the attacker. Researchers have shown in the past that flips to
page tables can cause privilege escalation and bypass security measures placed
by modern operating systems. It also has been shown that similar attacks also
work with memory chips used inside solid-state disks. Besides flipping bits in
page tables, researchers showed that when flipping bits in executables, the
behaviour of programs changes. With such changes, it is possible for an attacker
to abuse programs for privilege escalation or authentication-check bypasses.

We present a way of testing binaries for possible execution-path changes applied
by bitflips. In our work, we give a pre-defined outcome for a program and then
search for all single bitflips which would cause the executable to behave in the
desired way. Whereas previous work looked only at op-codes in machine
code, we scan the entire memory provided by the binary. In addition to the
executable, we also check all dynamically loaded libraries used by the program.
We show results for searched bitflips in the \texttt{sudo} program, for
bypassing user privilege checks, allowing super-user permissions without knowing
the password, and bitflips in the \texttt{nginx} web server, for bypassing HTTP
authentication, allowing an attacker to load files which he was not allowed to
access. In addition to searching for bitflips in static files, we also looked at
possible other attack vectors for Rowhammer. We show that when bitflips are
applied during the runtime of cryptographic calculations, these could break
assumptions made by the communicating parties and could even allow key leakage.
We apply bitflips to the implementation of AES-GCM in the OpenSSL library and
show how Rowhammer can be used to cause nonce-misuse.

With our work, we want to increase the awareness of Rowhammer and show how
software security is affected by bitflips. We call on to all vendors of hardware
to not forget to keep their systems secure and do not put competition and
performance ahead of security, which would harm their users.
\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{Kurzfassung}
\begin{otherlanguage}{ngerman}
\chapter*{Kurzfassung}
\label{cha:kurzfassung}

Angriffe auf die Mikroarchitektur zielen meistens auf Fehler von
Hardwareherstellern ab. Attacken auf Caches nutzen hierbei ein gewolltes
Verhalten des Systems aus. Populäre CPU Lücken wie Meltdown und Spectre machen
sich jedoch Fehler im Design der Hardware zu Nutzen, welche vom Hersteller
entweder durch Mikrocode Updates oder einem Tausch der CPU berichtigt werden
müssen. Rowhammer macht sich einen Designfehler in DRAM Chips zu Nutze.
Hersteller dieser Speicherchips wurden im Laufe der Zeit zu immer besserer
Performance und kleineren Chips gezwungen, hierbei wurde aber die
Qualitätskontrolle vernachlässigt. Mit speziellen Reihenfolgen von
Speicherzugriffen ist es möglich Interferenzen zu erzeugen welche eine
Ladungsänderung in benachbarten Speicherzellen verursacht, diese wechseln
dadurch ihren logischen Zustand.

Rowhammer Angriffe machen sich dieses Verhalten zu Nutze und zielen damit auf
Speicherbereiche ab, welche durch eine Änderung dem Angreifer einen Vorteil
verschaffen. Forscher haben in der Vergangenheit gezeigt, dass es möglich ist
mit Bitflips in Page Tables Privilegien einen Super-Users zu bekommen. Ebenso
wurde gezeigt, dass ähnliche Angriffe auch auf Speicherchips in
Solid-State-Disks möglich sind. Neben Flips in Page Tables wurde auch gezeigt,
dass Änderungen in ausführbaren Files Folgen auf das Verhalten des Programms
haben, sie ändern dieses durch das Wechseln eines einzigen Bits. Dies kann zum
Beispiel dazu führen dass Berechtigungsüberprüfungen umgangen werden.

Wir zeigen eine Möglichkeit Programme auf solche Verhaltensänderungen durch
Bitflips zu testen. Wir geben hier ein gewünschtes Verhalten vor und suchen
danach nach allen möglichen Bitflips welche das Verhalten in den gewünschten
Zustand ändern. Im Gegensatz zu früheren Arbeiten haben wir diesen Vorgang
automatisiert um eine größere Anzahl von Programmen abzudecken, zusätzlich
betrachten wir den gesamten Speicher in ausführbaren Files und dynamisch
geladenen Software-Bibliotheken. Wir zeigen die gefundenen Bitflips für das
\texttt{sudo} Programm, welche die Passwortabfrage umgehen und den
\texttt{nginx} Web-Server, wobei wir hier die HTTP-Authentication Checks
umgehen. Zusätzlich zur Untersuchung von statischen Files betrachten wir auch
die Auswirkung von Bitflips auf kryptographische Algorithmen. Hier untersuchen
wir, wie Rowhammer dazu genutzt werden kann um in der AES-GCM Implementation von
OpenSSL eine falsche Verwendung von Noncen zu verursachen.

Unsere Arbeit soll auf die Sicherheitsrisiken die durch Fehler wie Rowhammer
entstehen hinweisen und als Aufruf an die Hersteller von Hardware dienen, damit
diese nicht Wettbewerb und Performance über die Sicherheit ihrer Anwender
stellen.

\end{otherlanguage}

%\glsresetall %% all glossary entries should be used in long form (again)
%% vim:foldmethod=expr
%% vim:fde=getline(v\:lnum)=~'^%%%%\ .\\+'?'>1'\:'='
%%% Local Variables:
%%% mode: latex
%%% mode: auto-fill
%%% mode: flyspell
%%% eval: (ispell-change-dictionary "en_US")
%%% TeX-master: "main"
%%% End:
